### 2020-11-19 pldi21/abstract.tex
further tightening of abstract  typo fix

\begin{abstract}
    Although textual program manipulation is powerful, 
    some types of expressions are better represented and manipulated
    graphically.
    % Examples include expressions that compute 
    % colors,
    % animation parameters,
    % musical sequences,
    % tabular data, 
    % plot parameters,
    % GUI widgets, 
    % and various other domain-specific data structures.
    This paper introduces \emph{live literals}, or \emph{livelits}, 
    a mechanism that allows
    client programmers to fill holes of types like these 
    by directly manipulating a     
    provider-specified GUI embedded persistently into otherwise 
    symbolic code. 
    Uniquely, livelits are compositional: the GUI can itself contain
    spliced expressions, which can in turn contain other livelits. 
    Splices are typed and the system 
    ensures that the livelit treats each splice hygienically. 
    Livelits are also uniquely live: 
    they can offer {immediate feedback} about the run-time implications 
    of the programmer's choices 
    even when the spliced expressions mention bound variables, 
    because the system gathers closures associated with the 
    hole that the livelit is tasked with filling.
    This paper introduces livelits with case studies that exercise these novel capabilities. 
    We implement livelits in Hazel, 
    a live programming environment able to 
    typecheck and run programs with holes. 
    We then define a typed lambda calculus that captures the essence of livelits, which
    operate as live, graphical macros.     
    The macro expansion process has been mechanized in Agda.
\end{abstract}

>>>>

\begin{abstract}
    Although text editing is powerful, 
    some types of expressions are more naturally represented and manipulated
    graphically.
    % Examples include expressions that compute 
    % colors,
    % animation parameters,
    % musical sequences,
    % tabular data, 
    % plot parameters,
    % GUI widgets, 
    % and various other domain-specific data structures.
    This paper introduces \emph{live literals}, or \emph{livelits}, 
    which allow
    clients to fill holes of these types  
    by directly manipulating a     
    provider-defined GUI embedded persistently into otherwise 
    symbolic code. 
    Uniquely, livelits are compositional: the GUI can itself contain
    spliced expressions, which have full editor support and can in turn contain other livelits. 
    Splices are typed and the system 
    ensures that livelits treat splices hygienically. 
    Livelits are also uniquely live: 
    they can offer {immediate feedback} about the run-time implications 
    of the client's choices 
    even when splices mention bound variables, 
    because the system continuously gathers closures associated with the 
    hole that the livelit is tasked with filling.
    This paper introduces livelits with case studies that exercise these novel capabilities. 
    We implement livelits in Hazel, 
    a live programming environment able to 
    typecheck and run programs with holes. 
    We then define a typed lambda calculus that captures the essence of livelits as 
    live graphical macros.     
    The macro expansion process has been mechanized in Agda.
\end{abstract}

### 2020-11-19 pldi21/livelit-definitions.tex
further tightening of abstract  typo fix

Line 1 of Fig.~\ref{fig:color-impl} specifies the livelit's name, \li{\$color}, and its \emph{expansion type},
\li{Color}, which was discussed in Sec.~\ref{sec:livelit-expansion}.
The remainder of the definition is the livelit's implementation.
Livelits are implemented using a variation on the functional model-view-update
framework popularized by the Elm programming language\todo{cites}. We add a fourth component,
expansion generation. In addition, we add a simple monadic framework to provide the necessary  
interface between the livelit and the programming environment, discussed below, while retaining
a pure functional programming model.\todo{cite monads in Haskell}

\subsection{Model}\label{sec:model}
\todo{final check of line numbers}{}Line 2 of Fig.~\ref{fig:color-impl} specifies the livelit's \emph{model type},
here a labeled 4-tuple of \emph{splice references}, one for each of the four splices
that appear in the GUI in Fig.~\ref{fig:color}.
Models are persisted in the syntax tree of the program, so in practice,
 the system would check that the model type supports automatic serialization. 
 (Hazel is able to serialize all types because it is implemented as a 
 simple interpreter.)

The initial value of the model is specified starting on Line 4, 
which defines a command in the \li{UpdateMonad} that returns the 
initial model value after generating four new splice references
using the \li{new_splice} command:
\begin{lstlisting}[numbers=none]
  new_splice : (Typ, Option(Exp)) -> UpdateMonad(SpliceRef)
\end{lstlisting}
This command creates a fresh splice
of the given type and, optionally, with the given initial expression,
and returns a unique splice reference, which uniquely and abstractly identifies that splice.
Bolded types like \li{Typ}, \li{Exp}, \li{UpdateMonad}, and \li{SpliceRef}
are provided by the standard library. The \li{Typ} and \li{Exp} types 
encode the syntax of Hazel's types and expressions.
These two arguments are provided here in quasiquoted form, e.g. \li{`0`}\todo{cite something}. 
If no initial expression is provided, the splice is initialized with 
an empty hole.

The system checks that the provided type and initial content is valid 
assuming only the explicitly specified set of captured bindings on Line 3.
Here, we do not need to capture anything because \li{Int} is a built in type
and the initial content is closed. However, 
for splices of user-defined types, those types need to be explicitly captured.
Similarly, helper functions that are used in initial splices (and in the expansion,
discussed below) must be explicitly captured. 
This serves to ensure \emph{context independence} -- the livelit needs to make no 
assumptions about the typing context at use sites.
We use an explicit capture set, rather than implicitly capture of all bindings at the definition site, 
both to prevent inadvertent capture\todo{cite heathers paper}{} 
and to ensure that the capture set is part
of the interface of the livelit definition, which simplifies matters related to 
module export and determining paths to the captured bindings at use sites \cite{TLMs}.

\subsection{Action}
Line 11 defines the \li{Action} type for the \li{\$color} livelit, which 
specifies a single user-initiated action: clicking on a color using the user 
interface on the right side of Fig.~\ref{fig:color}. Actions are emitted
from event handlers (e.g. click handlers) defined in the computed \li{view}, 
and actions are consumed by the \li{update} function, causing a change in the model. 
Let us discuss each of these functions in turn.

@@

When the model is updated, a new view is 
computed. The system then performs a diff between the old and new view in order to 
efficiently perform the necessary imperative updates to the editor's visual state.
Changes to splices can also cause the view to be recomputed, because the view might 
need to evaluate the splices\todo{optimizaiton?}. The \li{update} function does not itself 
have the ability to request evaluation, because the model should not depend directly  
which closure the user has selected. Of course, the \li{view} might emit 
result-dependent actions when appropriate.

@@

The expansion can include splices, but we the system does not make the contents of a splice 
available directly. Instead, the \li{expand} must return a pair consisting of an encoded expression, of type 
\li{Exp}, with a list of \li{SpliceRef}s that the expansion depends on. 
We call the \li{Exp} the \emph{parameterized expansion}
 because it must take an argument for each listed \li{SpliceRef}s. 
 That argument will be the corresponding splice 
type, which was provided when the splice was initialized. 
 The return type of the parameterized expansion is the expansion type, here \li{Color}.

>>>>

Line 3 of Fig.~\ref{fig:color-impl} specifies the livelit's name, \li{\$color}, and its \emph{expansion type},
\li{Color}, which was discussed in Sec.~\ref{sec:livelit-expansion}.
The remainder of the definition is the livelit's implementation.
Livelits are implemented using a variation on the functional model-view-update
framework popularized by the Elm programming language \cite{ElmArchitecture}. We add a fourth component,
expansion generation. In addition, we add a simple monadic framework to provide the necessary  
interface between the livelit and the programming environment, discussed below, while retaining
a pure functional programming style.

\subsection{Model}\label{sec:model}
Line 6 of Fig.~\ref{fig:color-impl} specifies the livelit's \emph{model type},
here a labeled 4-tuple of \emph{splice references}, one for each of the four splices
that appear in the GUI in Fig.~\ref{fig:color}.
Models are persisted in the syntax tree of the program, so in practice,
 the system would check that the model type supports automatic serialization. 
 (Hazel is able to serialize all types because it is implemented as a 
 simple interpreter.)

The initial value of the model is specified starting on Line 7, 
which defines a command in the \li{UpdateMonad} that returns the 
initial model value after generating four new splice references
using the \li{new_splice} command:
\begin{lstlisting}[numbers=none]
new_splice : (Typ, Option(Exp)) -> UpdateMonad(SpliceRef)
\end{lstlisting}
This command creates a fresh splice
of the given type and, optionally, with the given initial expression,
and returns a unique splice reference, which uniquely and abstractly identifies that splice.
Bolded types like \li{Typ}, \li{Exp}, \li{UpdateMonad}, and \li{SpliceRef}
are provided by the standard library. The \li{Typ} and \li{Exp} types 
encode the syntax of Hazel's types and expressions.
These two arguments are provided here in quasiquoted form, e.g. \li{`0`} \cite{bawden1999quasiquotation}. 
If no initial expression is provided, the splice is initialized with 
an empty hole.

The system checks that the provided type and initial content is valid 
assuming only the explicitly specified set of captured bindings on Line 4.
Here, we do not need to capture any bindings because \li{Int} is a built in type
and the initial content is closed. However, 
for splices of user-defined types, those types need to be explicitly captured.
Similarly, helper functions that are used in initial splices (and in the expansion,
discussed below) must be explicitly captured. 
This serves to ensure \emph{context independence} -- the livelit needs to make no 
assumptions about the typing context at use sites.
We use an explicit capture set, rather than implicitly capture of all bindings at the definition site, 
both to prevent inadvertent capture 
and to ensure that the capture set is part
of the interface of the livelit definition, which simplifies matters related to 
module export and determining paths to the captured bindings at use sites \cite{TLMs}.
Formally, captured variables can be understood as parameters that have been 
immediately partially applied.

\subsection{Action}
Line 14 defines the \li{Action} type for the \li{\$color} livelit, which 
specifies a single user-initiated action: clicking on a color using the user 
interface on the right side of Fig.~\ref{fig:color}. Actions are emitted
from event handlers (e.g. click handlers) defined in the computed \li{view}, 
and actions are consumed by the \li{update} function, causing a change in the model. 
Let us discuss each of these functions in turn.

@@

When the model is updated, a new view is 
computed. The system then performs a diff between the old and new view in order to 
efficiently perform the necessary imperative updates to the editor's visual state.
Changes to splices can also cause the view to be recomputed, because the view might 
need to evaluate the splices. The \li{update} function does not itself 
have the ability to request evaluation, because the model should not depend directly  
which closure the user has selected. Of course, the \li{view} might emit 
result-dependent actions when appropriate.

@@

The expansion can include splices, but the system does not make the contents of a splice 
available directly. Instead, the \li{expand} must return a pair consisting of an encoded expression, of type 
\li{Exp}, with a list of \li{SpliceRef}s that the expansion depends on. 
We call the \li{Exp} the \emph{parameterized expansion}
 because it must take an argument for each listed \li{SpliceRef}s. 
 That argument will be the corresponding splice 
type, which was provided when the splice was initialized. 
 The return type of the parameterized expansion is the expansion type, here \li{Color}.
